Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis

### 2018-02-20 paper-ieee-trans.tex
homogeneize hyphens

\begin{abstract}
From the advent of Grid technology -- as the new paradigm of distributed
computing -- to the current days of Cloud computing models, the continuous need
of new tools and services to match the scientific community requirements has been
addressed in Europe by various projects dedicated to software development
and e-Infrastructure creation, operation and management.
% the creation of software development and e-Infrastructure coordination dedicated projects.
This work examines the evolution of the software engineering methodologies used in past and
present European Commission-funded projects to illustrate how the research on software
reliability has progressed in Europe over the last 15 years following the foundation of the
first project in 2001. In particular, we will summarize the techniques and procedures
applied to deliver quality software, highlighting the main challenges and barriers confronted
at the time, and how they were partially or totally overcome. The knowledge base established
throughout the years, sustained by the advances in the area of software engineering,
definitely conformed a significant breakthrough in the reliability of software delivered in the
European research arena. Recent software development projects funded by the European
Commission are good evidences of such insights, where the enforcement of Quality Assurance
procedures is present since the very early stages of the software development process.
\end{abstract}

@@

\IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
supporting an unified access to large-scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan-European research collaborations. One such example is the Worldwide
Large Hadron Collider (LHC) Computing Grid infrastructure, driven
by the particle physics scientific community, built to satisfy the computing needs
of processing, storing and analysis of the very large amount of data produced by the
LHC particle accelerator. A significant number of innovation projects and
technology transfer activities -- from the European Commission (EC)-funded 5th, 6th and 7th Framework Programmes,
and recently from the Horizon 2020 programme \cite{h2020} -- have as their main goal to
develop software solutions that provide a seamless and transparent access of researchers to distributed
computing and data resources.

Major developments were achieved throughout the years focused on providing new
functionalities, partially neglecting the adoption of software engineering methodologies, in the early days.
The released software was less tested, resulting in a high rate of
rollbacks and patches applied to production systems. Thereby, the stability and
reliability of the e-Infrastructures was often compromised, affecting
negatively their exploitation by the scientific communities.

Soon it became apparent the need for a better balance between the addition of new features
and the reliability of the released software. Consequently, a substantial part of the
software lifecycle would have to be devoted to the improvement of the quality of the software being
delivered. At this point, evolving software engineering practices were
considered and gradually adopted to define and implement quality procedures, the organization of cross-functional teams
and deployment of pilot testbeds.

Analysing the EC-funded software projects over the last decade,
one can observe a continuous increase in the prominence and robustness of software
testing and validation procedures. The
evolution of the software engineering methodologies (creation of the Agile
Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
with the parallel advancements in the information and communication technology
field (such as virtualization, the enabling technology for Cloud computing and
operating--system--level virtualization known as containerization) definitely motivated this trend. All of which
complemented by the emergence of automation and event-response tools.

In this paper, we outline a set of EC-funded projects
ranging from 2001 to nowadays, in which the authors were involved.
For each project, the challenges, the methodologies and
the achievements towards software reliability are described.

The reminder of this paper is
organized as follows. Section \ref{sec:ev} details the evolution of research on
software reliability in EC-projects. Section \ref{sec:ntsr} provides
information about new research trends in software reliability based upon the
experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section \ref{sec:con}.

\section{Evolution of research on software reliability in EC-projects}
\label{sec:ev}

@@

\begin{table}[!h]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{List of EC-projects}
\label{tab:eup}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
\begin{tabular}{p{1.6cm}p{1.5cm}p{3cm}l}
\hline
\hline
\\
Logo & Short Name & Long Name & Duration\\
\hline
\hline
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/datagrid}
\end{minipage}
    & DataGrid &
Research and Technological Development for an International Data Grid & 2001--2003\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egee}
\end{minipage}
     & EGEE I, II, III &
Enabling Grids for E-sciencE & 2004--2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/etics}
\end{minipage}
     & ETICS 1, 2 &
E--Infrastructure for Testing, Integration and Configuration of Software & 2006-2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/emi}
\end{minipage}
     & EMI &
Europeean Middleware Initiative & 2010--2013\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi-inspire}
\end{minipage}
     & EGI-Inspire &
Integrated Sustainable Pan-European Infrastructure for Researchers in Europe
 & 2010--2014\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi_engage}
\end{minipage}
     & EGI Engage &
Engaging the EGI Community towards an Open Science Commons
 & 2015--2017\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/indigo}
\end{minipage}
     & INDIGO--DataCloud &
INtegrating Distributed data Infrastructures for Global ExplOitation
 & 2015--2017\\
\hline
\hline
\end{tabular}
\end{table}

@@

The {\sl DataGrid} \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003)
main goal was to provide scientific communities, including physics, biology and
earth sciences, with intensive computing and large-scale dataset analysis capabilities.
The project brought together 21 academic and industry partners, from 15 different countries \cite{gagliardi}.
Grid was the emerging technology to be used in order to address their requirements,
thus the project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly
based on Globus middleware components and services \cite{globus}. The project was organized in 12 work packages, from
which 5 were devoted to software development and coordinated by an
Architecture Task Force, supervising the overall design and technical consistency
of the developments.

@@

\begin{table}[!h]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{Features on software reliability in EC-projects}
\label{tab:feat}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
\begin{adjustbox}{max width=0.5\textwidth}
\begin{tabular}{llllllll}
\hline
\hline
Feature & DataGrid & EGEEs & ETICSs & EMI & EGIs & INDIGO--DC\\
\hline
\hline
Architecture Task Force&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Communication Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Requirements Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Agile Methodologies&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Source code inspection&&&&$\surd$&&$\surd$\\
Build/Testing Management Procedure&$\surd$&$\surd$&$\surd$&$\surd$&&$\surd$\\
Software Product Metrics&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Quality Criteria Definition&&&$\surd$&$\surd$&$\surd$&$\surd$\\
%Manual Certification Procedure&$\surd$&$\surd$&&$\surd$&&\\
Automatic Certification&&&$\surd$&&$\surd$&$\surd$\\
Auto-generated Documentation&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
DevOps practices adoption (CI, CD)&&&&&&$\surd$\\
\hline
\hline
\end{tabular}
\end{adjustbox}
\end{table}

@@

The source code was available at a private, centralized version control system.
The code passed through a manual certification procedure at the time of the release.
This procedure aimed to improve the reliability of the software components by applying
acceptance criteria checks at the pre-release stage
\cite{egee:acceptance-criteria} i.e. integration, certification, pre--production and
production. Starting with {\sl EGEE--II}, the project adopted automation in the software
lifecycle process by leveraging the automatic build system for Grid middleware, that is
the {\sl ETICS} \cite{etics} solution.

@@

The E--Infrastructure for Testing, Integration and Configuration of Software
\cite{cordis:etics, cordis:etics2} ({\sl ETICS}, Jan 2006 -- Feb 2010) project aimed
at addressing the challenges in producing quality software in distributed,
collaborative projects such as {\sl EGEE} and its {\sl gLite} middleware. The framework
integrated different technologies and tools in order to provide automated configuration,
build and testing capabilities, as well as auto-generated documentation and
software metrics gathering such as Source Lines Of Code (SLOC), complexity and
number of defects/bugs \cite{etics}. The {\sl ETICS} framework was the first automated
service for delivering quality software products in distributed environments like
the Grids.

@@

In the following sections, we describe the three main pillars that sustain the project's strategy to attain reliability throughout the
software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
and metrics gathering definition}, provides guidelines to develop quality
software, validating each change in the code, and facilitating its adoption. The software lifecycle is continuously monitored
by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
process. A \textit{bottom-up adoption of DevOps principles} promotes automation to enable
the foregoing change-based development process and set up advanced delivery mechanisms.
The last key pillar is the \textit{two-factor validation} carried out over the project's pilot testbeds and
e-Infrastructure resource provider's early-adoption.

@@

\begin{itemize}
\item \textit{Code style}.
The ultimate goal for code style assessment is to improve the readability and reusability of the source code produced under the scope of the
project. Code style standards are enforced for every software component in the project stack. Community
most-adopted or \textit{de-facto standards} are initially recommended, however the development teams eventually
choose the set of guidelines to comply with.  Figure \ref{fig:fig_codestyle} highlights the code style standards and their popularity
in the {\sl INDIGO--DataCloud} software stack.

@@

\item \textit{Code review}.
Represents the last step in the change management pipeline, once the candidate change has
successfully passed through the testing methods described previously. It implies the human-based
revision of the proposed change to discuss its adequacy in terms of e.g. scope, objective fulfilment,
and documentation fulfilment. On approval, the candidate change is definitely merged into
the source code's production version. Human code reviewing is paramount in the software quality assessment, specially
important when the SQA testing requirements are fully automated. Secure code reviews are also 
done at this stage, assessing common vulnerabilities from inputs coming from automated linters
and manual dynamic application security testing.

@@

\item \textit{Automated Deployment}.
To lower the barriers of software adoption, the SQA criteria requires the automated
deployment of the products delivered as part of the catalogue. Automation in this context is
tackled using a configuration management tool. These tools allow to have an
\textit{Infrastructure as Code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as additional source of documentation
-- self-documenting code -- as they sequentially guide the component's deployment process in multiple
platforms. {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such
as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of
such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
the Ansible Galaxy portal \cite{indigo-galaxy}. Figure
\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
It shows an increase after the {\sl INDIGO--1} release and in the previous weeks of each
of the two major releases.
\end{itemize}

@@

\textit{Development metrics} are obtained programatically from several sources, namely GitHub
API and Jenkins service, and graphically displayed as GitHub pages using GrimoireLab
framework \cite{grimoirelab}. Per-component weekly reports, including the SQA requirement
fulfilment, are issued and individually discussed through the different communication channels
with the development teams. Issue tracking metrics and KPIs are an essential piece of information
of both feature addition and defect solving, useful to detect and fix misbehaviours while in the
development phase.

@@

\subsubsection{Continuous Integration (CI)}
\label{subsec:ci}
The {\sl INDIGO--DataCloud} project promoted the application of a CI scenario that enforces,
for any piece of produced software, the testing requirements defined in the SQA criteria,
as described in Section \ref{subsec:sqa}. Such environment requires an automation
ready--to--go infrastructure where the different services and technologies involved interact with each
other to trigger the source code validation pipeline for each candidate change. The
pipeline is mainly comprised by the code style validation, unit and functional testing
coverage. The CI pipeline is complemented with additional quality checks, such as
integration tests for the software components where automation of this type of testing
is applicable, or security linters for the code static analysis of suspicious
constructs that could lead to security risks. Metrics gathering also takes part of
the CI pipeline to have a per-change trend evolution of the common source code
related metrics, such as SLOC or the cyclomatic complexity.

@@

Jenkins \cite{jenkins}, the event-response CI system, and Docker container
provisioning to provide instant computing power needed for executing the pipeline jobs.
The project defined a source code contribution workflow, based on GitHub Pull Requests
(PRs), where each change automatically triggers in Jenkins the associated quality and
metric tracking checks on PR creation or update. The set of required checks contained
in the CI pipeline, report their exit status back to GitHub preventing the change to be
approved if they are not successfully executed. As each change is validated, the chances
of early detection of defects increases. Within this scenario, the cost of defect solving
is dramatically reduced and the reliability of the software solutions improved, as any
bug or design issue is likely to be detected and subsequently corrected in this phase.

@@

\subsubsection{Continuous Delivery}
As DevOps suggests, frequent releases positively affect the reliability of the software as
they involve fewer changes and usually are more correlated. The software updates of
{\sl INDIGO--DataCloud} products, taking place since the second major release, are passing through
a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
successful execution of the CI pipeline previously described. Consequently, CI-validated
software components are automatically delivered in online repositories without the need of
any human interaction but ultimately requiring expert supervision to validate the results.

{\sl INDIGO--DataCloud} software stack is delivered in form of Linux software packages (rpms, debs)
and Docker containers. The CD approach is different for each type of software packaging. \textit{Linux
package pipeline} intentionally delivers the packages produced in a pre-production -- preview --
branch. Before that, once the software packages have been created, they are uploaded to a
testing branch. The available automated deployment solution -- see Section \ref{subsec:sqa} -- checks
the installation and configuration of the component relying on the testing repository. On
successful completion, the packages are then automatically signed and subsequently moved to
the preview branch, from where they finally be made available in the production branch after
the release manager supervises the process.

@@

After more than a decade of experiences collected throughout the reviewed
EU-funded software development and e-Infrastructure management projects, it
becomes apparent that the European software produced for scientific purposes is
evolving to a more sustainable model where the quality and reliability of
software is being prioritized. The gradual adoption of software engineering insights, such as
DevOps practices or Agile methodologies, are leading the change. In this regard, the continuous evolution of the open--source
collaborative tools, extending their capabilities to bring along tight
integrations amongst other services, matched the requirements of such practices
and procedures, facilitating in a great manner their implementation in real scenarios.

Reliability of the software has been substantially improved by applying the SQA
criteria in the development stage. Consequently, each change in the source code
that is meant to be integrated in the production version, is validated according
to the SQA testing requirements. Thus, defects are detected and corrected early in the
software lifecycle, reducing the cost of fixing errors at later stages. Automation is
key in this process, allowing the implementation of a CI scenario that enforces
the SQA criteria compliance. The application of more advanced DevOps practices
resulted in the implementation of CD pipelines to extend automation to the
delivery of packages. Thereby, safe and frequent software releases are doable, reducing
the impact on production systems. In this regard, a pre-production validation,
through the preview testbeds or resource centers early adoption -- handled by the
Staged--rollout process -- alleviates the likelihood of discovering uncovered
issues at production.

Software lifecycle demands as well from a continuous measurement and analysis
process that permits timely reactions to discover and fix defects throughout the
development, release and post-release or maintenance stages. Conscientious metric
tracking and response sits at the core of an appropriate software maturity, as
ultimate reliability can only be achieved by enforcing and empowering the SQA
criteria through each stage of the software lifecycle.

@@

{\sl DataGrid - Research and Technological Development for an International Data Grid} project 
has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
FP5-IST - Programme for research, technological development and demonstration on a "User-friendly information society, 1998-2002" (IST--2000--25182E)

{\sl EGEE - Enabling grids For E-science} project
has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
 FP6-INFRASTRUCTURES - Research infrastructures: Specific programme for research, technological development and demonstration: "Structuring the European Research Area" under the Sixth Framework Programme 2002-2006, (INFSO--RI--508833)

{\sl EGEE - Enabling grids For E-science--II} project
 has received funding from the European Commission - Community Research and Development Information System (CORDIS) under 
 FP6-2005-INFRASTRUCTURES-7 programme, Topic - "INFRASTR-3 - Communication network development in conjunction with thematic priority 2 (Information Society Technologies) to establish a high-capacity and high-speed communications network for all researchers in Europe (GEANT)", (INFSO--RI--031688)

{\sl EGEE - Enabling grids For E-science--III} project
  has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
  FP7-INFRASTRUCTURES - Specific Programme "Capacities": Research infrastructures, Topic - "INFRA-2007-1.2-03 - e-Science Grid infrastructures", (INFSO--RI--222667)

@@

\bibitem{zhu}
L. Zhu and L. Bass and G. Champlin-Scharff, \emph{DevOps and Its Practices},
IEEE Software, vol. 33, no. 3, pp. 32--34, May-June 2016.

>>>>

\begin{abstract}
From the advent of Grid technology -- as the new paradigm of distributed
computing -- to the current days of Cloud computing models, the continuous need
of new tools and services to match the scientific community requirements has been
addressed in Europe by various projects dedicated to software development
and e--Infrastructure creation, operation and management.
% the creation of software development and e-Infrastructure coordination dedicated projects.
This work examines the evolution of the software engineering methodologies used in past and
present European Commission--funded projects to illustrate how the research on software
reliability has progressed in Europe over the last 15 years following the foundation of the
first project in 2001. In particular, we will summarize the techniques and procedures
applied to deliver quality software, highlighting the main challenges and barriers confronted
at the time, and how they were partially or totally overcome. The knowledge base established
throughout the years, sustained by the advances in the area of software engineering,
definitely conformed a significant breakthrough in the reliability of software delivered in the
European research arena. Recent software development projects funded by the European
Commission are good evidences of such insights, where the enforcement of Quality Assurance
procedures is present since the very early stages of the software development process.
\end{abstract}

@@

\IEEEPARstart{D}uring the last 15 years the provision of an European e--Infrastructure,
supporting an unified access to large--scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan--European research collaborations. One such example is the Worldwide
Large Hadron Collider (LHC) Computing Grid infrastructure, driven
by the particle physics scientific community, built to satisfy the computing needs
of processing, storing and analysis of the very large amount of data produced by the
LHC particle accelerator. A significant number of innovation projects and
technology transfer activities -- from the European Commission (EC)-funded 5th, 6th and 7th Framework Programmes,
and recently from the Horizon 2020 programme \cite{h2020} -- have as their main goal to
develop software solutions that provide a seamless and transparent access of researchers to distributed
computing and data resources.

Major developments were achieved throughout the years focused on providing new
functionalities, partially neglecting the adoption of software engineering methodologies, in the early days.
The released software was less tested, resulting in a high rate of
rollbacks and patches applied to production systems. Thereby, the stability and
reliability of the e--Infrastructures was often compromised, affecting
negatively their exploitation by the scientific communities.

Soon it became apparent the need for a better balance between the addition of new features
and the reliability of the released software. Consequently, a substantial part of the
software lifecycle would have to be devoted to the improvement of the quality of the software being
delivered. At this point, evolving software engineering practices were
considered and gradually adopted to define and implement quality procedures, the organization of cross--functional teams
and deployment of pilot testbeds.

Analysing the EC--funded software projects over the last decade,
one can observe a continuous increase in the prominence and robustness of software
testing and validation procedures. The
evolution of the software engineering methodologies (creation of the Agile
Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
with the parallel advancements in the information and communication technology
field (such as virtualization, the enabling technology for Cloud computing and
operating--system--level virtualization known as containerization) definitely motivated this trend. All of which
complemented by the emergence of automation and event--response tools.

In this paper, we outline a set of EC--funded projects
ranging from 2001 to nowadays, in which the authors were involved.
For each project, the challenges, the methodologies and
the achievements towards software reliability are described.

The reminder of this paper is
organized as follows. Section \ref{sec:ev} details the evolution of research on
software reliability in EC--projects. Section \ref{sec:ntsr} provides
information about new research trends in software reliability based upon the
experiences from the recent {\sl INDIGO--DataCloud} project (INDIGO--DC). Conclusions are drawn in Section \ref{sec:con}.

\section{Evolution of research on software reliability in EC--projects}
\label{sec:ev}

@@

\begin{table}[!h]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{List of EC--projects}
\label{tab:eup}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
\begin{tabular}{p{1.6cm}p{1.5cm}p{3cm}l}
\hline
\hline
\\
Logo & Short Name & Long Name & Duration\\
\hline
\hline
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/datagrid}
\end{minipage}
    & DataGrid &
Research and Technological Development for an International Data Grid & 2001--2003\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egee}
\end{minipage}
     & EGEE I, II, III &
Enabling Grids for E--sciencE & 2004--2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/etics}
\end{minipage}
     & ETICS 1, 2 &
E--Infrastructure for Testing, Integration and Configuration of Software & 2006-2010\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/emi}
\end{minipage}
     & EMI &
Europeean Middleware Initiative & 2010--2013\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi-inspire}
\end{minipage}
     & EGI--Inspire &
Integrated Sustainable Pan--European Infrastructure for Researchers in Europe
 & 2010--2014\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/egi_engage}
\end{minipage}
     & EGI Engage &
Engaging the EGI Community towards an Open Science Commons
 & 2015--2017\\
\begin{minipage}{.3\textwidth}
\includegraphics[width=15mm,height=7.5mm]{images/indigo}
\end{minipage}
     & INDIGO--DataCloud &
INtegrating Distributed data Infrastructures for Global ExplOitation
 & 2015--2017\\
\hline
\hline
\end{tabular}
\end{table}

@@

The {\sl DataGrid} \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003)
main goal was to provide scientific communities, including physics, biology and
earth sciences, with intensive computing and large--scale dataset analysis capabilities.
The project brought together 21 academic and industry partners, from 15 different countries \cite{gagliardi}.
Grid was the emerging technology to be used in order to address their requirements,
thus the project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly
based on Globus middleware components and services \cite{globus}. The project was organized in 12 work packages, from
which 5 were devoted to software development and coordinated by an
Architecture Task Force, supervising the overall design and technical consistency
of the developments.

@@

\begin{table}[!h]
%% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
% if using array.sty, it might be a good idea to tweak the value of
% \extrarowheight as needed to properly center the text within the cells
\caption{Features on software reliability in EC--projects}
\label{tab:feat}
\centering
%% Some packages, such as MDW tools, offer better commands for making tables
%% than the plain LaTeX2e tabular which is used here.
\begin{adjustbox}{max width=0.5\textwidth}
\begin{tabular}{llllllll}
\hline
\hline
Feature & DataGrid & EGEEs & ETICSs & EMI & EGIs & INDIGO--DC\\
\hline
\hline
Architecture Task Force&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Communication Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Requirements Handling&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Agile Methodologies&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Source code inspection&&&&$\surd$&&$\surd$\\
Build/Testing Management Procedure&$\surd$&$\surd$&$\surd$&$\surd$&&$\surd$\\
Software Product Metrics&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
Quality Criteria Definition&&&$\surd$&$\surd$&$\surd$&$\surd$\\
%Manual Certification Procedure&$\surd$&$\surd$&&$\surd$&&\\
Automatic Certification&&&$\surd$&&$\surd$&$\surd$\\
Auto--generated Documentation&&$\surd$&$\surd$&$\surd$&$\surd$&$\surd$\\
DevOps practices adoption (CI, CD)&&&&&&$\surd$\\
\hline
\hline
\end{tabular}
\end{adjustbox}
\end{table}

@@

The source code was available at a private, centralized version control system.
The code passed through a manual certification procedure at the time of the release.
This procedure aimed to improve the reliability of the software components by applying
acceptance criteria checks at the pre--release stage
\cite{egee:acceptance-criteria} i.e. integration, certification, pre--production and
production. Starting with {\sl EGEE--II}, the project adopted automation in the software
lifecycle process by leveraging the automatic build system for Grid middleware, that is
the {\sl ETICS} \cite{etics} solution.

@@

The E--Infrastructure for Testing, Integration and Configuration of Software
\cite{cordis:etics, cordis:etics2} ({\sl ETICS}, Jan 2006 -- Feb 2010) project aimed
at addressing the challenges in producing quality software in distributed,
collaborative projects such as {\sl EGEE} and its {\sl gLite} middleware. The framework
integrated different technologies and tools in order to provide automated configuration,
build and testing capabilities, as well as auto--generated documentation and
software metrics gathering such as Source Lines Of Code (SLOC), complexity and
number of defects/bugs \cite{etics}. The {\sl ETICS} framework was the first automated
service for delivering quality software products in distributed environments like
the Grids.

@@

In the following sections, we describe the three main pillars that sustain the project's strategy to attain reliability throughout the
software lifecycle. A solid \textit{Software Quality Assurance (SQA) criteria
and metrics gathering definition}, provides guidelines to develop quality
software, validating each change in the code, and facilitating its adoption. The software lifecycle is continuously monitored
by a set of metrics, allowing promptly reactions to detected malfunctions throughout the
process. A \textit{bottom--up adoption of DevOps principles} promotes automation to enable
the foregoing change--based development process and set up advanced delivery mechanisms.
The last key pillar is the \textit{two--factor validation} carried out over the project's pilot testbeds and
e--Infrastructure resource provider's early--adoption.

@@

\begin{itemize}
\item \textit{Code style}.
The ultimate goal for code style assessment is to improve the readability and reusability of the source code produced under the scope of the
project. Code style standards are enforced for every software component in the project stack. Community
most--adopted or \textit{de--facto standards} are initially recommended, however the development teams eventually
choose the set of guidelines to comply with.  Figure \ref{fig:fig_codestyle} highlights the code style standards and their popularity
in the {\sl INDIGO--DataCloud} software stack.

@@

\item \textit{Code review}.
Represents the last step in the change management pipeline, once the candidate change has
successfully passed through the testing methods described previously. It implies the human--based
revision of the proposed change to discuss its adequacy in terms of e.g. scope, objective fulfilment,
and documentation completeness. On approval, the candidate change is definitely merged into
the source code's production version. Human code reviewing is paramount in the software quality assessment, specially
important when the SQA testing requirements are fully automated. Secure code reviews are also 
done at this stage, assessing common vulnerabilities from inputs coming from automated linters
and manual dynamic application security testing.

@@

\item \textit{Automated Deployment}.
To lower the barriers of software adoption, the SQA criteria requires the automated
deployment of the products delivered as part of the catalogue. Automation in this context is
tackled using a configuration management tool. These tools allow to have an
\textit{Infrastructure as Code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as additional source of documentation
-- self--documenting code -- as they sequentially guide the component's deployment process in multiple
platforms. {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such
as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of
such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
the Ansible Galaxy portal \cite{indigo-galaxy}. Figure
\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
It shows an increase after the {\sl INDIGO--1} release and in the previous weeks of each
of the two major releases.
\end{itemize}

@@

\textit{Development metrics} are obtained programmatically from several sources, namely GitHub
API and Jenkins service, and graphically displayed as GitHub pages using GrimoireLab
framework \cite{grimoirelab}. Per--component weekly reports, including the SQA requirement
fulfilment, are issued and individually discussed through the different communication channels
with the development teams. Issue tracking metrics and KPIs are an essential piece of information
of both feature addition and defect solving, useful to detect and fix misbehaviours while in the
development phase.

@@

\subsubsection{Continuous Integration (CI)}
\label{subsec:ci}
The {\sl INDIGO--DataCloud} project promoted the application of a CI scenario that enforces,
for any piece of produced software, the testing requirements defined in the SQA criteria,
as described in Section \ref{subsec:sqa}. Such environment requires an automation
ready--to--go infrastructure where the different services and technologies involved interact with each
other to trigger the source code validation pipeline for each candidate change. The
pipeline is mainly comprised by the code style validation, unit and functional testing
coverage. The CI pipeline is complemented with additional quality checks, such as
integration tests for the software components where automation of this type of testing
is applicable, or security linters for the code static analysis of suspicious
constructs that could lead to security risks. Metrics gathering also takes part of
the CI pipeline to have a per--change trend evolution of the common source code
related metrics, such as SLOC or the cyclomatic complexity.

@@

Jenkins \cite{jenkins}, the event--response CI system, and Docker container
provisioning to provide instant computing power needed for executing the pipeline jobs.
The project defined a source code contribution workflow, based on GitHub Pull Requests
(PRs), where each change automatically triggers in Jenkins the associated quality and
metric tracking checks on PR creation or update. The set of required checks contained
in the CI pipeline, report their exit status back to GitHub preventing the change to be
approved if they are not successfully executed. As each change is validated, the chances
of early detection of defects increases. Within this scenario, the cost of defect solving
is dramatically reduced and the reliability of the software solutions improved, as any
bug or design issue is likely to be detected and subsequently corrected in this phase.

@@

\subsubsection{Continuous Delivery}
As DevOps suggests, frequent releases positively affect the reliability of the software as
they involve fewer changes and usually are more correlated. The software updates of
{\sl INDIGO--DataCloud} products, taking place since the second major release, are passing through
a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
successful execution of the CI pipeline previously described. Consequently, CI--validated
software components are automatically delivered in online repositories without the need of
any human interaction but ultimately requiring expert supervision to validate the results.

{\sl INDIGO--DataCloud} software stack is delivered in form of Linux software packages (rpms, debs)
and Docker containers. The CD approach is different for each type of software packaging. \textit{Linux
package pipeline} intentionally delivers the packages produced in a pre--production -- preview --
branch. Before that, once the software packages have been created, they are uploaded to a
testing branch. The available automated deployment solution -- see Section \ref{subsec:sqa} -- checks
the installation and configuration of the component relying on the testing repository. On
successful completion, the packages are then automatically signed and subsequently moved to
the preview branch, from where they finally be made available in the production branch after
the release manager supervises the process.

@@

After more than a decade of experiences collected throughout the reviewed
EU--funded software development and e--Infrastructure management projects, it
becomes apparent that the European software produced for scientific purposes is
evolving to a more sustainable model where the quality and reliability of
software is being prioritized. The gradual adoption of software engineering insights, such as
DevOps practices or Agile methodologies, are leading the change. In this regard, the continuous evolution of the open--source
collaborative tools, extending their capabilities to bring along tight
integrations amongst other services, matched the requirements of such practices
and procedures, facilitating in a great manner their implementation in real scenarios.

Reliability of the software has been substantially improved by applying the SQA
criteria in the development stage. Consequently, each change in the source code
that is meant to be integrated in the production version, is validated according
to the SQA testing requirements. Thus, defects are detected and corrected early in the
software lifecycle, reducing the cost of fixing errors at later stages. Automation is
key in this process, allowing the implementation of a CI scenario that enforces
the SQA criteria compliance. The application of more advanced DevOps practices
resulted in the implementation of CD pipelines to extend automation to the
delivery of packages. Thereby, safe and frequent software releases are doable, reducing
the impact on production systems. In this regard, a pre--production validation,
through the preview testbeds or resource centers early adoption -- handled by the
Staged--rollout process -- alleviates the likelihood of discovering uncovered
issues at production.

Software lifecycle demands as well from a continuous measurement and analysis
process that permits timely reactions to discover and fix defects throughout the
development, release and post--release or maintenance stages. Conscientious metric
tracking and response sits at the core of an appropriate software maturity, as
ultimate reliability can only be achieved by enforcing and empowering the SQA
criteria through each stage of the software lifecycle.

@@

{\sl DataGrid - Research and Technological Development for an International Data Grid} project 
has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
FP5-IST - Programme for research, technological development and demonstration on a "User--friendly information society, 1998-2002" (IST--2000--25182E)

{\sl EGEE - Enabling grids For E--science} project
has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
 FP6-INFRASTRUCTURES - Research infrastructures: Specific programme for research, technological development and demonstration: "Structuring the European Research Area" under the Sixth Framework Programme 2002-2006, (INFSO--RI--508833)

{\sl EGEE - Enabling grids For E--science--II} project
 has received funding from the European Commission - Community Research and Development Information System (CORDIS) under 
 FP6-2005-INFRASTRUCTURES-7 programme, Topic - "INFRASTR-3 - Communication network development in conjunction with thematic priority 2 (Information Society Technologies) to establish a high--capacity and high--speed communications network for all researchers in Europe (GEANT)", (INFSO--RI--031688)

{\sl EGEE - Enabling grids For E--science--III} project
  has received funding from the European Commission - Community Research and Development Information System (CORDIS) under
  FP7-INFRASTRUCTURES - Specific Programme "Capacities": Research infrastructures, Topic - "INFRA-2007-1.2-03 - e--Science Grid infrastructures", (INFSO--RI--222667)

@@

\bibitem{zhu}
L. Zhu and L. Bass and G. Champlin--Scharff, \emph{DevOps and Its Practices},
IEEE Software, vol. 33, no. 3, pp. 32--34, May--June 2016.
